Evolution of pre-copulatory and post-copulatory strategies of inbreeding avoidance

### 2017-04-05 ms_pre_post.Rmd
Final tweaks before sending to potential coauthors

---
title: Evolution of pre-copulatory and post-copulatory strategies of inbreeding avoidance
  and associated polyandry
author: ''
date: "23 March 2017"
output:
  pdf_document:
    fig_caption: yes
  html_document: default
  word_document:
    fig_caption: yes
    pandoc_args:
    - --csl
    - evolution.csl
    reference_docx: docx_template.docx
header-includes:
- \usepackage{natbib}
- \usepackage{lineno}
- \linenumbers
- \bibliographystyle{amnatnat}
linestretch: 2
csl: evolution.csl
bibliography: ms_pre_post.bib
---

@@

Evolution of different reproductive strategies cannot be presumed to occur in isolation from each other in real-world populations. Rather, such evolution is likely to involve multiple interacting phenotypes resulting in the potential for feedback and degeneracy (i.e., functional redundance) among phenotypes under selection. For example, strong inbreeding depression in offspring fitness is widely hypothesised to drive selection for inbreeding avoidance, which could consequently drive selection for polyandry [e.g., @Zeh1997; @Jennions2000; @Tregenza2002]. Evolution of inbreeding avoidance can be driven by selection on both pre-copulatory and post-copulatory mechanisms, so simultaneous evolution of each phenotype is likely to be defined by degeneracy, in addition to any direct costs on phenotypes. We used individual-based modelling to explore evolution of pre-copulatory and post-copulatory inbreeding avoidance and polyandry given (1) hard constraints on evolution of post-copulatory inbreeding avoidance, (2) asymmetric costs of pre-copulatory and post-copulatory inbreeding avoidance, and (3) evolution of one inbreeding avoidance phenotype given the pre-existence the other. We thereby highlight fundamental but theoretically under-developed relationships between evolution of female mating frequency and pre-copulatory versus post-copulatory inbreeding avoidance. We present our model and simulation results as a tool for thinking more clearly about reproductive strategies, not for making accurate quantitative predictions in any specific system. While the logic of our model can be usefully applied to constructing hypotheses within and across empirical systems, accurate quantitative prediction would require additional detail and the incorporation of empirical data for model parameterisation.

@@

Evolution of interacting pre-copulatory and post-copulatory inbreeding avoidance occurred in our model, but was affected by the joint evolution of polyandry and the existence of asymmetric costs. One cost free strategy of inbreeding avoidance precluded another more costly strategy from persisting in a focal population by modifying its effect on inbreeding avoidance (Figure 3). Consequently, long-term persistence of both pre-copulatory and post-copulatory inbreeding avoidance might not be expected in populations given sufficient cost asymmetry. Since direct costs strongly affected the evolution of the focal phenotype(s), quantifying costs is necessary for predicting evolution of inbreeding strategies. Quantifying direct costs of phenotypes associated with polyandry and mate choice is empirically challenging [@Pomiankowski1987; @Kokko2003]. @Pomiankowski1987 categorised four types of costs that are relevant to mate choice, which are also relevant for polyandry, and include costs of elevated risks of predation or disease transmission, and costs incurred through time or energy expenditure. Another type of cost relevant for polyandry and inbreeding strategy is risk of male harm caused by sexual conflict [e.g., @Arnqvist2005a; @Parker2006]. We assumed that males were passive in mating encounters with females, but inbreeding theory predicts that males should benefit from a higher tolerance of inbreeding than females, leading to sexual conflict over inbreeding [@Parker1979; @Parker2006; @Kokko2006; @Duthie2015a]. Here we interpreted a cost of polyandy as an elevated risk of predation as might be incurred while searching for or courting mates [e.g., @Rowe1988; @Rowe1994], a cost of pre-copulatory inbreeding strategy as a lost of time [i.e., risk of not finding a mate in time due to choosiness; e.g., @Kokko2013], and a cost of post-copulatory inbreeding strategy as depletion of energy. While these costs thereby reflect empirical observations in some populations and are therefore biologically realistic, other types of costs have also been observed and could therefore be considered in future models. For example, increased risk of disease transmission has been observed for highly polyandrous females [@Roberts2015a], but such a cost would more realistically be applied to the number of realised mates a female has rather than her tendency for engaging in polyandry. Different types of costs might affect selection both directly and indirectly through interactions between reproductive strategies, thereby affecting evolution of entire reproductive systems. Evolution of individual phenotypes therefore needs to be considered in the broader context of potentially interacting and phenotypes and their costs.

In our model, existence of adaptive pre-copulatory inbreeding avoidance precluded evolution of polyandry and, in turn, evolution of post-copulatory inbreeding avoidance (Figure 3C,D). However, existence of adaptive post-copulatory inbreeding avoidance did not preclude evolution of pre-copulatory inbreeding avoidance (Figure 3A,B). In wild populations, it is unlikely that pre-copulatory and post-copulatory inbreeding avoidance will evolve simultaneously from an ancestral population in which females mate and assign paternity randomly. Rather, for example, initial evolution of pre-copulatory inbreeding avoidance is much more likely to evolve in the absence of evolving post-copulatory inbreeding avoidance due to constraints, or in the presence of already evolved post-copulatory inbreeding avoidance. When framing hypotheses for existence of post-copulatory inbreeding avoidance and polyandry, it might therefore be necessary to consider whether or not inbreeding avoidance already occurs through pre-copulatory mate choice. Additionally, the opportunity for post-copulatory inbreeding avoidance will also depend on the degree to which females are polyandrous. For species in which pre-copulatory inbreeding avoidance occurs and polyandry is uncommon [@Lihoreau2007; @Metzger2010; @Metzger2010a], evolution of post-copulatory inbreeding avoidance is highly unlikely even if such a strategy incurs little direct cost.

@@

Post-copulatory inbreeding avoidance cannot be realised if females are not polyandrous, and is most effective for highly polyandrous females that can choose amongst multiple mates and therefore bias fertilisation accordingly. In contrast, pre-copulatory inbreeding avoidance mechanisms are most critical for females that mate only once and therefore have no post-copulatory opportunity to avoiding inbreeding. This theory is borne out in our simulation results, as selection for, and consequent evolution of, post-copulatory inbreeding avoidance was negligible in populations where polyandry did not evolve, resulting in high drift of allele values over generations due to the inability of females to express post-copulatory choice (e.g., Figure 4C,D). Evolution of pre-copulatory inbreeding avoidance was also typically slightly stronger when polyandry did not evolve (e.g., Figure 3A vs 3B; see also Supporting Information p. S1). In addition to polyandry causing evolution of post-copulatory inbreeding avoidance, polyandry might also covary positively with post-copulatory inbreeding avoidance due to the feed back that post-copulatory inbreeding avoidance has on facilitating evolution of polyandry itself, as observed in our model (Figure 1). It would therefore be interesting to test the hypothesis that, across taxa, occurrence of post-copulatory inbreeding avoidance covaries positively, and pre-copulatory inbreeding avoidance covaries negatively, with degree of polyandry. To test this hypothesis, more work is needed to identify the degree to which females of different species engage in polyandry and the degree to which females express both pre-copulatory and post-copulatory inbreeding avoidance.

Degeneracy occurs at nearly all biological scales [@Edelman2001], including complex systems affecting organismal development [e.g., @Nowak1997], adaptation [@Whitacre2010; @Whitacre2010a], and cognition [@Price2002; @Park2013], as well as population [@Atamas2009], community [@Suraci2017], and ecosystem [e.g., @Levin2008] dynamics. In our model, degeneracy occurred through overlaps in how different reproductive strategies caused adaptive inbreeding avoidance. In general, degeneracy might increase biological robustness by fine-tuning degenerate phenotypes to different local environments [@Gardner2006; @Whitacre2010a]. For example, degeneracy might ensure successful inbreeding avoidance through pre-copulatory or post-copulatory mechanisms where or when avoidance through the other mechanism is ineffective (e.g., due to sexual conflict affecting mate choice or injury affecting fertilisation). However, evolution of one inbreeding avoidance mechanism might also weaken selection on the other by modifying the latter's impact on total realised inbreeding avoidance [*sensu* evolution of genetic redundancy; see @Nowak1997]. The relevance of degeneracy with respect to such reproductive strategies therefore requires further theoretical development, which could result in new empirical predictions and conceptual synthesis across biological scales.

@@

![Mean allele values underlying tendency for polyandry (red), pre-copulatory inbreeding strategy (blue), and post-copulatory inbreeding strategy (black) when (A and B) costly pre-copulatory inbreeding strategy ($c_{M} = 0.02$) can evolve and post-copulatory inbreeding strategy is (A) fixed for random fertilisation or (B) can also evolve, and when (C and D) costly post-copulatory inbreeding avoidance ($c_{F} = 0.02$) can evolve and pre-copulatory inbreeding strategy is (C) fixed for random mating or (D) can also evolve. Mean allele values (solid lines) and associated standard errors (shading) are calculated across all individuals within a population over 40000 generations across 40 replicate populations. Negative mean allele values indicate strategies of inbreeding avoidance or tendency for monandry, and positive values indicate strategies of inbreeding preference or tendency for polyandry. In all panels, polyandry is cost free.](figures/rel_costs.png)

>>>>

---
title: Evolution of pre-copulatory and post-copulatory strategies of inbreeding avoidance
  and associated polyandry
author: ''
date: "5 April 2017"
output:
  word_document:
    fig_caption: yes
    pandoc_args:
    - --csl
    - evolution.csl
    reference_docx: docx_template.docx
  pdf_document:
    fig_caption: yes
  html_document: default
header-includes:
- \usepackage{natbib}
- \usepackage{lineno}
- \linenumbers
- \bibliographystyle{amnatnat}
linestretch: 2
csl: evolution.csl
bibliography: ms_pre_post.bib
---

@@

Evolution of different reproductive strategies cannot be presumed to occur in isolation from each other in real-world populations. Rather, such evolution is likely to involve multiple interacting phenotypes resulting in the potential for feedback and degeneracy (i.e., functional redundance) among phenotypes under selection. For example, strong inbreeding depression in offspring fitness is widely hypothesised to drive selection for inbreeding avoidance, which could consequently drive selection for polyandry [e.g., @Zeh1997; @Jennions2000; @Tregenza2002]. Evolution of inbreeding avoidance can be driven by selection on both pre-copulatory and post-copulatory mechanisms, so simultaneous evolution of each phenotype is likely to be defined by degeneracy, in addition to any direct costs on phenotypes. We used individual-based modelling to explore evolution of pre-copulatory and post-copulatory inbreeding strategy and polyandry given (1) hard constraints on evolution of post-copulatory inbreeding strategy, (2) asymmetric costs of pre-copulatory and post-copulatory inbreeding strategy, and (3) evolution of one inbreeding strategy phenotype given the pre-existence of the other. We thereby highlight fundamental but theoretically under-developed relationships between evolution of female mating frequency and pre-copulatory versus post-copulatory inbreeding strategy. We present our model and simulation results as a tool for thinking more clearly about the dynamics of simultaneously evolving reproductive strategies, not for making accurate quantitative predictions for any specific system. While the logic of our model can be usefully applied to constructing hypotheses within and across empirical systems, accurate quantitative prediction would require additional detail and the incorporation of empirical data for model parameterisation.

@@

Evolution of interacting pre-copulatory and post-copulatory inbreeding avoidance occurred in our model, but was affected by the joint evolution of polyandry and the existence of asymmetric costs. One cost free strategy of inbreeding avoidance precluded another more costly strategy from persisting in a focal population by modifying its effect on inbreeding avoidance (Figure 3). Consequently, long-term persistence of both pre-copulatory and post-copulatory inbreeding avoidance might not be expected in populations given sufficient cost asymmetry. Since direct costs strongly affected the evolution of the focal phenotype(s), quantifying costs is necessary for predicting evolution of inbreeding strategies. Quantifying direct costs of phenotypes associated with polyandry and mate choice is empirically challenging [@Pomiankowski1987; @Kokko2003]. @Pomiankowski1987 categorised four types of costs that are relevant to mate choice, which are also relevant for polyandry, and include costs of elevated risks of predation or disease transmission, and costs incurred through time or energy expenditure. Another relevant cost is risk of male harm caused by sexual conflict [e.g., @Arnqvist2005a; @Parker2006]. We assumed that males were passive in mating encounters with females, but inbreeding theory predicts that males should benefit from a higher tolerance of inbreeding than females, leading to sexual conflict over inbreeding [@Parker1979; @Parker2006; @Kokko2006; @Duthie2015a]. Here we interpreted a cost of polyandy as an elevated risk of predation as might be incurred while searching for or courting mates [e.g., @Rowe1988; @Rowe1994], a cost of pre-copulatory inbreeding strategy as a lost of time [i.e., risk of not finding a mate in time due to choosiness; e.g., @Kokko2013], and a cost of post-copulatory inbreeding strategy as an up-front depletion of energy. While these costs reflect empirical observations in some populations and are therefore biologically realistic, other types of costs have also been observed and could be considered in future models tailored to specific systems. For example, increased risk of disease transmission has been observed for highly polyandrous females [@Roberts2015a], and such a cost would more realistically be applied to the number of realised mates a female has rather than her tendency for engaging in polyandry. Different types of costs might affect selection both directly and indirectly through interactions between reproductive strategies, thereby affecting evolution of entire reproductive systems. Evolution of individual phenotypes therefore needs to be considered in the broader context of potentially interacting and phenotypes and their costs.

In our model, existence of adaptive pre-copulatory inbreeding avoidance precluded evolution of polyandry and, in turn, evolution of post-copulatory inbreeding avoidance (Figure 3C,D). However, existence of adaptive post-copulatory inbreeding avoidance did not preclude evolution of pre-copulatory inbreeding avoidance (Figure 3A,B). In wild populations, it is unlikely that pre-copulatory and post-copulatory inbreeding avoidance will evolve simultaneously from an ancestral population in which females mate and assign paternity randomly. Rather, the timing of the invasion of adaptive inbreeding avoidance phenotypes will differ, so the initial evolution of one inbreeding strategy will likely occur in the absence of the other, or where selection for and subsequent evolution of the other strategy has already occurred. When framing hypotheses for existence of post-copulatory inbreeding avoidance and polyandry, it might therefore be necessary to consider whether or not inbreeding avoidance already occurs through pre-copulatory mate choice. Additionally, the opportunity for post-copulatory inbreeding avoidance will also depend on the degree to which females are polyandrous. For species in which pre-copulatory inbreeding avoidance occurs and polyandry is uncommon [@Lihoreau2007; @Metzger2010; @Metzger2010a], evolution of post-copulatory inbreeding avoidance is highly unlikely even if such a strategy incurs little direct cost.

@@

Post-copulatory inbreeding avoidance cannot be realised if females are not polyandrous, and is most effective for highly polyandrous females that can choose amongst multiple mates and therefore bias fertilisation accordingly. In contrast, pre-copulatory inbreeding avoidance mechanisms are most critical for females that mate only once and therefore have no post-copulatory opportunity to avoiding inbreeding. This theory is borne out in our simulation results, as selection for, and consequent evolution of, post-copulatory inbreeding avoidance was negligible in populations where polyandry did not evolve, resulting in high drift of allele values over generations due to the inability of females to express post-copulatory choice (e.g., Figure 4C,D). Evolution of pre-copulatory inbreeding avoidance was also typically slightly stronger when polyandry did not evolve (e.g., Figure 3A vs 3B; see also Supporting Information Figure S9). In addition to polyandry causing evolution of post-copulatory inbreeding avoidance, polyandry might also covary positively with post-copulatory inbreeding avoidance due to the feed back that post-copulatory inbreeding avoidance has on facilitating evolution of polyandry itself, as observed in our model (Figure 1). It would therefore be interesting to test the hypothesis that, across taxa, occurrence of post-copulatory inbreeding avoidance covaries positively, and pre-copulatory inbreeding avoidance covaries negatively, with the degree of polyandry. To test this hypothesis, more work is needed to quantify the degree to which females of different species engage in polyandry and the degree to which females express both pre-copulatory and post-copulatory inbreeding avoidance.

Degeneracy occurs at nearly all biological scales [@Edelman2001], including complex systems affecting organismal development [e.g., @Nowak1997], adaptation [@Whitacre2010; @Whitacre2010a], and cognition [@Price2002; @Park2013], as well as population [@Atamas2009], community [@Suraci2017], and ecosystem [e.g., @Levin2008] dynamics. In our model, degeneracy occurred through overlaps in how different reproductive strategies caused adaptive inbreeding avoidance. In general, degeneracy might increase biological robustness by fine-tuning degenerate phenotypes to different local environments [@Gardner2006; @Whitacre2010a]. For example, degeneracy might ensure successful inbreeding avoidance through either pre-copulatory or post-copulatory mechanisms when avoidance through the other mechanism is ineffective (e.g., due to sexual conflict affecting mate choice or injury affecting fertilisation). However, evolution of one inbreeding avoidance mechanism might also weaken selection on the other by modifying the latter's impact on total realised inbreeding avoidance [*sensu* evolution of genetic redundancy; see @Nowak1997]. The relevance of degeneracy with respect to such reproductive strategies therefore requires further theoretical development, which could result in new empirical predictions and conceptual synthesis across biological scales.

@@

![Mean allele values underlying tendency for polyandry (red), pre-copulatory inbreeding strategy (blue), and post-copulatory inbreeding strategy (black) when (A and B) costly pre-copulatory inbreeding strategy ($c_{M} = 0.02$) can evolve and post-copulatory inbreeding strategy is (A) fixed for random fertilisation or (B) can also evolve, and when (C and D) costly post-copulatory inbreeding strategy ($c_{F} = 0.02$) can evolve and pre-copulatory inbreeding strategy is (C) fixed for random mating or (D) can also evolve. Mean allele values (solid lines) and associated standard errors (shading) are calculated across all individuals within a population over 40000 generations across 40 replicate populations. Negative mean allele values indicate strategies of inbreeding avoidance or tendency for monandry, and positive values indicate strategies of inbreeding preference or tendency for polyandry. In all panels, polyandry is cost free.](figures/rel_costs.png)
